---
title: "limbBudAnalysis_geneNetworksWGCNA"
output: html_document
---



```{r}
source("https://bioconductor.org/biocLite.R")
biocLite()
biocLite("WGCNA")
# Flags not allowing multi-threading in wgcna, prompts use of this command to allow:
library(WGCNA)
allowWGCNAThreads()
```

To start this new type of analysis I will run through a tutorial posted by the original authors. 
https://horvath.genetics.ucla.edu/html/CoexpressionNetwork/Rpackages/WGCNA/Tutorials/
This site includes R scripts to download, which I have done and followed. I will be working with these scripts in the background. 


Select data:
====================================================================================
Tutorial selected 3,600 probes out of a larger set of 20,000 on the basis of "most variable and most connected(??) probes".
- Since this ran easily enough on my machine, I will take 5,000 probes.
Selected probesets based on:
- top5000variance
- top5000StDv
- top5000mad (median absolute deviation)
- all statistically significant probes
- all biologically and statistically significant probes (log2FC > 1.5)

Load up the dds object from the start of limbBudAnalysis_markdown.Rmd
```{r}
# variance
top5000VarianceNames <- names(sort(apply(assay(dds),1,var),decreasing=T))[1:5000]

```

